The Weakly Nonlinear Magnetorotational Instability

### 2016-09-02 paper/widegap_mri_paper.tex
taylor couette --> TC

The diversity of astrophysical systems which may be MRI unstable yields an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI--and even its very existence--may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, and so forth. Multiphysics numerical simulations of such systems is currently an area of intense focus, enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study \citep[e.g.][among many others]{Fleming:2003fs,Bai:2011cm, Flock:2013,Suzuki:2014vh}. Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. For example, the magnetic Prandtl number $\Pm = \nu/\eta \sim 10^{-8}$ in protoplanetary disks \citet[e.g.][]{OishiMacLow:2011} and $\sim 10^{-6}$ in liquid metal experiments \citep[e.g.][]{Goodman:2002ix}. Such extreme ratios of viscosity to resistivity far exceed current computational resources. However, we can construct asymptotic approximations valid for $\Pm << 1$ using analytic methods. Analytic methods can also play a powerful role in elucidating the mechanisms responsible for MRI saturation. For instance, analytical approaches have revealed the mechanism that likely governs saturation in the ``shearing box'' approximation. The shearing box is an oft-invoked local approximation in which a section of a disk is represented by solving the MHD equations in a rotating, Cartesian box with a linearized background shear, subject to shear periodic boundary conditions in the radial direction. The shearing box is a convenient computational framework allowing extreme resolution for local MRI studies and has been extended to include vertical stratification and a wide variety of diffusive effects. However, while the MRI is a local instability, there are a number of important problems that require a global treatment. Perhaps most importantly, linear evolution in the shearing box is dominated by channel modes, particularly when a net vertical magnetic field threads the box. These linear modes are exact solutions to the \textit{nonlinear} local MRI equations. The shearing box MRI system avoids runaway growth by a secondary instability of the channel modes themselves (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-represented in the shearing box (e.g. \citei{Latter:2015}). Thus while the shearing box may accurately approximate many features of the global MRI, the saturation mechanism may not be among them.

@@

In this paper, we develop an analytic, global theory for the MRI in a Taylor-Couette flow. This system precludes channel modes, allowing us to develop an understanding of MRI saturation in their absence. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation (e.g. \citei{Knobloch:2005ba}, \citei{Umurhan:2007hs}). The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI (e.g. \citei{Ebrahimi:2009ey}). The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge is not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which will allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI in a spherical Couette flow, but most likely detected unrelated MHD instabilities instead (\citei{Hollerbach:2009ig}, \citei{Gissinger:2011td}). Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium Taylor-Couette flow with an axial magnetic field (\citei{Ji:2001kd}). There has been some theoretical work designed to complement the Princeton MRI experiment involving direct numerical simulation of the experimental conditions, much of it focused on the specific challenges in identifying MRI signatures despite apparatus-driven spurious flows (e.g. \citei{Gissinger:2012gc}). The vertical endcaps on a laboratory MRI apparatus drive meridional flows which both inhibit the excitement of MRI and obscure its detection. The Princeton MRI experiment employs split, independently rotating endcaps to mitigate these flows (\citei{Schartman:2009df}). Our work assumes an infinite vertical domain, an idealization that is theoretically expedient but experimentally impractical. However, our setup is designed to be an accurate treatment of the radial dimension of the flow in a Taylor Couette apparatus like the one used in the Princeton MRI experiment.

@@

Because we include curvature terms, our treatment also allows us to study the helical magnetorotational instability (HMRI). The HMRI is an overstability in which the background magnetic field is helical, $\mathbf{B} = B_0 (\xi \mathbf{\hat{\theta}} + \mathbf{\hat{z}})$ \citet{Hollerbach:2005tr}. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the $\Pm$ becomes very small. However the rotation profiles needed to excite HMRI may be steeper than Keplerian, depending on the boundary conditions, and so its role in astrophysical disks is currently a matter of debate \citep{Liu:2006,Rudiger:2007,Kirillov:2013}. Regardless of its astrophysical role, the HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}).

@@

The astrophysical magnetorotational instability operates in accretion disks and in stellar interiors, environments where fluid rotation is strongly regulated by gravity. In accretion disks, differential rotation is imposed gravitationally by a central body, so the rotation profile is forced to be Keplerian. Clearly a gravitationally enforced Keplerian flow is inaccessible to laboratory study, so differential rotation is created by rotating an inner cylinder faster than an outer cylinder (a Taylor-Couette setup). For a nonideal fluid subject to no-slip boundary conditions, the base flow is

@@

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection (\citei{Newell:1969wr}), and hydrodynamic Taylor Couette flow (e.g. \citei{Recktenwald:1993}). We emphasize that this is a model equation, valid only near marginality (\citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ is in general complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (see \citei{Aranson:2002} for a thorough review).

@@

We examine the behavior of the wide gap MRI system as a function of $\Pm$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $\Pm$, as well as the GLE linear coefficient $b$ and the diffusion coefficient $d$. The GLE coefficients are remarkably insensitive to $\Pm$. From Equation \ref{eq:gle} it is readily apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, so we conclude that the saturation amplitude of the MRI is only very weakly dependent on $\Pm$. Note that because $\reym$ is essentially constant as a function of $\Pm$, the saturation amplitude is equivalently insensitive to $\reye^{-1}$. This is in stark contrast to the narrow gap behavior of the system. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $\Pm^{2/3}$. They find that this amplitude dependence is driven by the $\Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in Taylor Couette flow (\citei{Goodman:2002ix}). %Why does the $\Pm$ scaling of the boundary layer width not translate to a stronger $\Pm$ scaling for the saturation amplitude in the wide gap case? The boundary layer in the wide gap case is strongly localized at the walls, i.e. $r_1$ and $r_2$. 
Figure \ref{fig:wide_narrow_N31B} shows the structure of the third-order nonlinear term $N_{31}^{(A)}$ as a function of $\Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$, and thus the scaling of the saturation amplitude because of the insensitivity of $b$ to $\Pm$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs}, \citei{Clark:2016} for the narrow gap equations). Clearly, the width of the boundary layers scales with $\Pm$ in both the wide and narrow gap MRI. This translates to a steeper saturation amplitude $\Pm$ dependence in the wide gap case.%However, in the narrow gap case this scaling extends prominently into the center of the channel, whereas for the wide gap case the bulk of the flow is relatively unaffected by the boundary layer scaling.

@@

\section{Discussion}
\label{sec:discussion}
Our work should be placed in the broader context of emergent pattern formation in physical systems. The real Ginzburg-Landau equation derived here governs the slow-parameter evolution of the standard MRI close to threshold. The GLE arises in a number of other physical systems, and in each case it is a consequence not of the particular physics at hand, but of the underlying symmetries in the problem. Here we make a phenomenological comparison to two other systems that give rise to a GLE. The first and perhaps most famous is Rayleigh-B\'enard convection, in which a fluid between two plates is heated from below (\citei{Newell:1969wr}). If we take the plane of the fluid to be infinite in the horizontal plane, the system is initially translationally symmetric. %When the system eventually becomes unstable to convective heat transfer it undergoes a symmetry breaking, allowing the 
At the onset of convection the system undergoes a symmetry breaking, forming rolls, or convection cells, which break the horizontal translational invariance. Analogously, the standard MRI system considered here is initially vertically translationally symmetric, because we idealize the Taylor-Couette device as an infinitely long cylinder. The MRI breaks this symmetry, forming cells along the vertical length of the domain. Just as Rayleigh-B\'enard cells transport heat vertically, the MRI cells transport angular momentum horizontally. The symmetry breaking of each of these systems is described near onset by the real GLE.

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (\citei{Parker:2013hy}). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (\citei{Johansen:2009uj}, \citei{Kunz:2013}). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the application of this model system to realistic astrophysical disks. Our model is most directly relevant to Taylor-Couette flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

>>>>

The diversity of astrophysical systems which may be MRI unstable yields an enormous parameter space to be explored. In protoplanetary disks, for example, the behavior and evolution of the MRI--and even its very existence--may change drastically depending on the properties of the magnetic field, the disk composition, disk geometry, and so forth. Multiphysics numerical simulations of such systems is currently an area of intense focus, enabling the study of nonideal MHD effects, disk stratification, nonequilibrium chemistry, and other complex physics that does not lend itself easily to analytic study \citep[e.g.][among many others]{Fleming:2003fs,Bai:2011cm, Flock:2013,Suzuki:2014vh}. Still, computational costs inevitably constrain numerical approaches. MRI saturation is a complicated nonlinear problem which may depend on the assumptions and approximations adopted by simulations in nonobvious ways. For example, the magnetic Prandtl number $\Pm = \nu/\eta \sim 10^{-8}$ in protoplanetary disks \citet[e.g.][]{OishiMacLow:2011} and $\sim 10^{-6}$ in liquid metal experiments \citep[e.g.][]{Goodman:2002ix}. Such extreme ratios of viscosity to resistivity far exceed current computational resources. However, we can construct asymptotic approximations valid for $\Pm << 1$ using analytic methods. Analytic methods can also play a powerful role in elucidating the mechanisms responsible for MRI saturation. For instance, analytical approaches have revealed the mechanism that likely governs saturation in the ``shearing box'' approximation. The shearing box is an oft-invoked local approximation in which a section of a disk is represented by solving the MHD equations in a rotating, Cartesian box with a linearized background shear, subject to shear periodic boundary conditions in the radial direction. The shearing box is a convenient computational framework allowing extreme resolution for local MRI studies and has been extended to include vertical stratification and a wide variety of diffusive effects. However, while the MRI is a local instability, there are a number of important problems that require a global treatment. Perhaps most importantly, linear evolution in the shearing box is dominated by channel modes, particularly when a net vertical magnetic field threads the box. These linear modes are exact solutions to the \textit{nonlinear} local MRI equations. The shearing box MRI system avoids runaway growth by a secondary instability of the channel modes themselves (\citei{Goodman:1994ul}, \citei{Pessah:2010ic}). The growth of parasitic modes provides a saturation avenue for channel mode-dominated flows, yet this is unlikely to be the dominant saturation mechanism in laboratory experiments or astrophysical disks, as channel modes are artificially over-represented in the shearing box \citep{Latter:2015}. Thus while the shearing box may accurately approximate many features of the global MRI, the saturation mechanism may not be among them.

@@

In this paper, we develop a weakly nonlinear, global theory for the MRI in a Taylor-Couette (TC) geometry. This system precludes channel modes, allowing us to develop an understanding of MRI saturation in their absence. A number of saturation mechanisms have been proposed for the MRI which do not rely on channel modes dominating the flow. The MRI feeds off of the free energy from differential rotation, and so a modification of the background shear may cause saturation \citep{Knobloch:2005ba, Umurhan:2007hs}. The MRI may transfer its free energy into the magnetic field, and saturate when the field is too strong to be susceptible to the MRI \citep{Ebrahimi:2009ey}. The MRI may saturate differently depending on the particular parameter regime under investigation, and so our challenge is not only in identifying possible saturation mechanisms, but in understanding how and when each applies in different astrophysical environments.

Our investigation is astrophysically motivated, but we also intend our theory to be relevant to laboratory experiments. Several experimental efforts are attempting to observe the MRI in the laboratory, which will allow the study of a crucial astrophysical phenomenon in a controlled setting. Unfortunately, detection of the MRI has so far proven elusive. \citei{Sisan:2004ig} claimed to detect the MRI in a spherical Couette flow, but most likely detected unrelated MHD instabilities instead \citep{Hollerbach:2009ig,Gissinger:2011td}. Most relevant to our work is the Princeton Plasma Physics Laboratory (PPPL) MRI Experiment, a liquid gallium TC flow with an axial magnetic field \citep{Ji:2001kd}. There has been some theoretical work designed to complement the Princeton MRI experiment involving direct numerical simulation of the experimental conditions, much of it focused on the specific challenges in identifying MRI signatures despite apparatus-driven spurious flows (e.g. \citei{Gissinger:2012gc}). The vertical endcaps on a laboratory MRI apparatus drive meridional flows which both inhibit the excitement of MRI and obscure its detection. The Princeton MRI experiment employs split, independently rotating endcaps to mitigate these flows (\citei{Schartman:2009df}). Our work assumes an infinite vertical domain, an idealization that is theoretically expedient but experimentally impractical. Such an approach changes the symmetry properties of the solution significantly, and in the much better studied hydrodynamic case this leads to significant differences even for TC devices with very large aspect ratios \citep{Lopez:2005}. However, our setup is designed to be an accurate treatment of the radial dimension of the flow in a TC apparatus like the one used in the Princeton MRI experiment.

@@

Because we include curvature terms, our treatment also allows us to study the helical magnetorotational instability (HMRI). The HMRI is an overstability in which the background magnetic field is helical, $\mathbf{B} = B_0 (\xi r/r_0 \mathbf{\hat{\phi}} + \mathbf{\hat{z}})$ \citet{Hollerbach:2005tr}. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the $\Pm$ becomes very small. However the rotation profiles needed to excite HMRI may be steeper than Keplerian, depending on the boundary conditions, and so its role in astrophysical disks is currently a matter of debate \citep{Liu:2006,Rudiger:2007,Kirillov:2013}. Regardless of its astrophysical role, the HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}).

@@

The astrophysical magnetorotational instability operates in accretion disks and in stellar interiors, environments where fluid rotation is strongly regulated by gravity. In accretion disks, differential rotation is imposed gravitationally by a central body, so the rotation profile is forced to be Keplerian. Clearly a gravitationally enforced Keplerian flow is inaccessible to laboratory study, so differential rotation is created by rotating an inner cylinder faster than an outer cylinder (a TC setup). For a nonideal fluid subject to no-slip boundary conditions, the base flow is

@@

a Ginzburg-Landau equation (GLE). The GLE governs the weakly nonlinear amplitude behavior in a wide range of physical systems, including the narrow gap MRI (\citei{Umurhan:2007hs}), Rayleigh-B\'enard convection (\citei{Newell:1969wr}), and hydrodynamic TC flow (e.g. \citei{Recktenwald:1993}). We emphasize that this is a model equation, valid only near marginality (\citei{Cross:1993el}). The dynamics of the GLE are determined by its coefficients, which are in turn determined by the linear eigenfunctions and nonlinear vectors plotted in Figures \ref{fig:linear_eigenfunctions} and \ref{fig:N2_N3}. Equation \ref{eq:gle} contains three coefficients: $b$, which determines the linear growth rate of the system, $d$, a diffusion coefficient, and $c$, the coefficient of the nonlinear term. When all of the coefficients of Equation \ref{eq:gle} are real, this is known as the real GLE, although the amplitude $\alpha$ is in general complex. The real GLE is subject to several well-studied instabilities, including the Ekhaus and Zig-Zag instabilities. When the coefficients are complex, we have the complex GLE, a source of even richer phase dynamics than its real counterpart (see \citei{Aranson:2002} for a thorough review).

@@

We examine the behavior of the wide gap MRI system as a function of $\Pm$. Figure \ref{fig:coefficients} shows the critical parameters $k_c$ and $\reym$ as a function of $\Pm$, as well as the GLE linear coefficient $b$ and the diffusion coefficient $d$. The GLE coefficients are remarkably insensitive to $\Pm$. From Equation \ref{eq:gle} it is readily apparent that the asymptotic saturation amplitude is $\alpha_{s} = \pm \sqrt{b/c}$, so we conclude that the saturation amplitude of the MRI is only very weakly dependent on $\Pm$. Note that because $\reym$ is essentially constant as a function of $\Pm$, the saturation amplitude is equivalently insensitive to $\reye^{-1}$. This is in stark contrast to the narrow gap behavior of the system. For these same boundary conditions, \citeauthor{Umurhan:2007hs} (\citeyear{Umurhan:2007hs}) find that the narrow gap saturation amplitude scales as $\Pm^{2/3}$. They find that this amplitude dependence is driven by the $\Pm^{1/3}$ dependence of the linear boundary layer. Boundary layer analysis similarly reveals a $\nu^{1/3}$ dependence for the radial extent of the boundary layer in TC flow (\citei{Goodman:2002ix}). %Why does the $\Pm$ scaling of the boundary layer width not translate to a stronger $\Pm$ scaling for the saturation amplitude in the wide gap case? The boundary layer in the wide gap case is strongly localized at the walls, i.e. $r_1$ and $r_2$. 
Figure \ref{fig:wide_narrow_N31B} shows the structure of the third-order nonlinear term $N_{31}^{(A)}$ as a function of $\Pm$ for both the narrow and wide gap standard MRI. $N_{31}$ is the vector that determines the GLE coefficient $c$, and thus the scaling of the saturation amplitude because of the insensitivity of $b$ to $\Pm$ (see Appendix \ref{app:basic_equations} for the wide gap case, and \citei{Umurhan:2007hs}, \citei{Clark:2016} for the narrow gap equations). Clearly, the width of the boundary layers scales with $\Pm$ in both the wide and narrow gap MRI. This translates to a steeper saturation amplitude $\Pm$ dependence in the wide gap case.%However, in the narrow gap case this scaling extends prominently into the center of the channel, whereas for the wide gap case the bulk of the flow is relatively unaffected by the boundary layer scaling.

@@

\section{Discussion}
\label{sec:discussion}
Our work should be placed in the broader context of emergent pattern formation in physical systems. The real Ginzburg-Landau equation derived here governs the slow-parameter evolution of the standard MRI close to threshold. The GLE arises in a number of other physical systems, and in each case it is a consequence not of the particular physics at hand, but of the underlying symmetries in the problem. Here we make a phenomenological comparison to two other systems that give rise to a GLE. The first and perhaps most famous is Rayleigh-B\'enard convection, in which a fluid between two plates is heated from below (\citei{Newell:1969wr}). If we take the plane of the fluid to be infinite in the horizontal plane, the system is initially translationally symmetric. %When the system eventually becomes unstable to convective heat transfer it undergoes a symmetry breaking, allowing the 
At the onset of convection the system undergoes a symmetry breaking, forming rolls, or convection cells, which break the horizontal translational invariance. Analogously, the standard MRI system considered here is initially vertically translationally symmetric, because we idealize the TC device as an infinitely long cylinder. The MRI breaks this symmetry, forming cells along the vertical length of the domain. Just as Rayleigh-B\'enard cells transport heat vertically, the MRI cells transport angular momentum horizontally. The symmetry breaking of each of these systems is described near onset by the real GLE.

A real GLE has also been found to describe the formation of zonal flows out of magnetized turbulence in a model system (\citei{Parker:2013hy}). Zonal flows are axisymmetric structures, large-scale and long-lived, which form spontaneously out of turbulence. They have recently been observed in some numerical studies of the MRI, and have generated considerable interest for their possible role in planet formation in protoplanetary disks (\citei{Johansen:2009uj}, \citei{Kunz:2013}). The present work is of course an idealized geometry, and we make no attempt to model a realistic protoplanetary disk environment. However, it is worth noting that the GLE we derive implies that axisymmetric, large-scale, long-lived structures are a generic feature of the MRI in the weakly nonlinear regime. This work provides a mathematical description of the MRI as a pattern-forming process, but much remains to be understood, particularly involving the application of this model system to realistic astrophysical disks. Our model is most directly relevant to TC flows, and we emphasize that laboratory MRI experiments stand poised to observe the MRI-driven pattern formation predicted here.

@@

\bibitem[Lopez \& Marques(2005)]{Lopez:2005} 
Lopez, J.~M., \& Marques, F.\ 2005, Physica D, 211, 168
